---
title: "DEXSeq"
author: "Marika Oksanen"
date: "2022-09-15"
output: html_document
---

This markdown includes DEXSeq analysis between HNRNPU deficiency models and their respective neurotypical controls at D0, D5 and D28. 

Please note that HNRNPUdel is occasionally referred as ASD12BII

```{r}
library(DEXSeq)
library(tidyverse)
```

## DEXSeq 

#### D0

Load count files to R environment:

```{r}
countFiles = list.files("countfiles/HNRNPUdelD0", pattern=".txt$", full.names=TRUE)
```

```{r}
flattenedFile = list.files(pattern="gff$", full.names=TRUE)
```

```{r}
sample_info <- read.delim("Sample_info.txt", header=TRUE)
```

```{r}
sample_info <- sample_info[c(1:5,9:13),]
sample_info <- separate(sample_info, "User.ID", into = c("sample_number", "cell_line", "timepoint"), sep = "_", remove = FALSE )
ReplicateNumber <- c(1:5, 1:5)
rownames(sample_info) <- paste(sample_info$cell_line, ReplicateNumber, sep = "_")
names(sample_info)[names(sample_info) == "cell_line"] <- "condition"
```

Preparing DEXSeq data set:

```{r}
dxd_D0 = DEXSeqDataSetFromHTSeq(
   countFiles,
   sampleData=sample_info,
   design= ~ sample + exon + condition:exon,
   flattenedfile=flattenedFile )
```

dispersion estimation:
```{r}
dxd_D0 <- estimateSizeFactors(dxd_D0)
dxd_D0 <- estimateDispersions(dxd_D0)
plotDispEsts(dxd_D0)
```

Performing dexseq with default settings:
```{r}
dxr_D0 = DEXSeq(dxd_D0)
```



#### D5

Load D5 count files to R environment:

```{r}
countFiles = list.files("countfiles/HNRNPUdelD5", pattern=".txt$", full.names=TRUE)
```

```{r}
sample_info <- read.delim("Sample_info.txt", header=TRUE)
```

```{r}
sample_info <- sample_info[c(6:8,14:17,35:37),]
sample_info <- separate(sample_info, "User.ID", into = c("sample_number", "cell_line", "timepoint"), sep = "_", remove = FALSE )
sample_info <- sample_info[rev(order(sample_info$cell_line)),]
ReplicateNumber <- c(1:5, 1:5)
rownames(sample_info) <- paste(sample_info$cell_line, ReplicateNumber, sep = "_")
names(sample_info)[names(sample_info) == "cell_line"] <- "condition"
```

Preparing DEXSeq data set:

```{r}
dxd_D5 = DEXSeqDataSetFromHTSeq(
   countFiles,
   sampleData=sample_info,
   design= ~ sample + exon + condition:exon,
   flattenedfile=flattenedFile )
```

dispersion estimation:
```{r}
dxd_D5 <- estimateSizeFactors(dxd_D5)
dxd_D5 <- estimateDispersions(dxd_D5)
plotDispEsts(dxd_D5)
```

Performing dexseq with default settings:
```{r}
dxr_D5 = DEXSeq(dxd_D5)
```


#### D28

Load count files to R environment:

```{r}
countFiles = list.files("countfiles/HNRNPUdelD28", pattern=".txt$", full.names=TRUE)
```

```{r}
sample_info <- read.delim("Sample_infoD28.txt", header=TRUE)
```

```{r}
sample_info <- sample_info[44:49,]
sample_info <- separate(sample_info, "User.ID", into = c("sample_number", "cell_line", "timepoint"), sep = "_", remove = FALSE )
sample_info <- sample_info[rev(order(sample_info$cell_line)),]
ReplicateNumber <- c(1:3, 1:3)
rownames(sample_info) <- paste(sample_info$cell_line, ReplicateNumber, sep = "_")
names(sample_info)[names(sample_info) == "cell_line"] <- "condition"
```

Preparing DEXSeq data set:
```{r}
dxd_D28 = DEXSeqDataSetFromHTSeq(
   countFiles,
   sampleData=sample_info,
   design= ~ sample + exon + condition:exon,
   flattenedfile=flattenedFile )
```

dispersion estimation:
```{r}
dxd_D28 <- estimateSizeFactors(dxd_D28)
dxd_D28 <- estimateDispersions(dxd_D28)
plotDispEsts(dxd_D28)
```

Performing dexseq with default settings:
```{r}
dxr_D28 = DEXSeq(dxd_D28)
```


#### siRNA D0

Load count files to R environment:

```{r}
countFiles = list.files("countfiles/siHNRNPUD0", pattern=".txt$", full.names=TRUE)
```

```{r}
sample_info <- read.delim("Sample_info.txt", header=TRUE)
```

```{r}
sample_info <- sample_info[c(18:20,26:30,38,39),]
sample_info <- separate(sample_info, "User.ID", into = c("sample_number", "cell_line", "treatment", "timepoint"), sep = "_", remove = FALSE )
sample_info <- sample_info[order(sample_info$treatment, decreasing = TRUE),]
ReplicateNumber <- c(1:5, 1:5)
rownames(sample_info) <- paste(sample_info$treatment, ReplicateNumber, sep = "_")
names(sample_info)[names(sample_info) == "treatment"] <- "condition"
```

Preparing DEXSeq data set:

```{r}
dxd_siRNAD0 = DEXSeqDataSetFromHTSeq(
   countFiles,
   sampleData=sample_info,
   design= ~ sample + exon + condition:exon,
   flattenedfile=flattenedFile )
```

dispersion estimation:
```{r}
dxd_siRNAD0 <- estimateSizeFactors(dxd_siRNAD0)
dxd_siRNAD0 <- estimateDispersions(dxd_siRNAD0)
plotDispEsts(dxd_siRNAD0)
```

Performing dexseq with default settings:
```{r}
dxr_siRNAD0 = DEXSeq(dxd_siRNAD0)
```

#### siRNA D5

Load count files to R environment:
```{r}
countFiles = list.files("countfiles/siHNRNPUD5", pattern=".txt$", full.names=TRUE)
```

```{r}
sample_info <- read.delim("Sample_info.txt", header=TRUE)
```

```{r}
sample_info <- sample_info[c(21:25, 31:34, 40),]
sample_info <- separate(sample_info, "User.ID", into = c("sample_number", "cell_line", "treatment", "timepoint"), sep = "_", remove = FALSE )
ReplicateNumber <- c(1:5, 1:5)
rownames(sample_info) <- paste(sample_info$treatment, ReplicateNumber, sep = "_")
names(sample_info)[names(sample_info) == "treatment"] <- "condition"
```

Preparing DEXSeq data set:

```{r}
dxd_siRNAD5 = DEXSeqDataSetFromHTSeq(
   countFiles,
   sampleData=sample_info,
   design= ~ sample + exon + condition:exon,
   flattenedfile=flattenedFile )
```

dispersion estimation:
```{r}
dxd_siRNAD5 <- estimateSizeFactors(dxd_siRNAD5)
dxd_siRNAD5 <- estimateDispersions(dxd_siRNAD5)
plotDispEsts(dxd_siRNAD5)
```

Performing dexseq with default settings:
```{r}
dxr_siRNAD5 = DEXSeq(dxd_siRNAD5)
```


#### siRNA D28

Load count files to R environment:

```{r}
countFiles = list.files("countfiles/siHNRNPUD28", pattern=".txt$", full.names=TRUE)
```

```{r}
sample_info <- read.delim("Sample_info_siRNAD28.txt", header=TRUE)
```

```{r}
sample_info <- sample_info[order(sample_info$StatusHNRNPU, decreasing = FALSE),]
rownames(sample_info) <- paste(sample_info$StatusHNRNPU, sample_info$ReplicateNumber, sep = "_")
names(sample_info)[names(sample_info) == "StatusHNRNPU"] <- "condition"
sample_info <- sample_info[order(rownames(sample_info)),]
```

Preparing DEXSeq data set:

```{r}
dxd_siRNAD28 = DEXSeqDataSetFromHTSeq(
   countFiles,
   sampleData=sample_info,
   design= ~ sample + exon + condition:exon,
   flattenedfile=flattenedFile )
```

dispersion estimation:
```{r}
dxd_siRNAD28 <- estimateSizeFactors(dxd_siRNAD28)
dxd_siRNAD28 <- estimateDispersions(dxd_siRNAD28)
plotDispEsts(dxd_siRNAD28)
```

Performing dexseq with default settings:
```{r}
dxr_siRNAD28 = DEXSeq(dxd_siRNAD28)
```
