---
title: "Gene_Enrichment_Analysis"
author: "F Mastropasqua"
date: "10/5/2022"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("dplyr")
```

```{r}
SFARI_sel_list <- read.csv("./SFARI_release_072022_list.txt", header = TRUE)
DD_list <- read.csv("./NDD_list.txt", header = FALSE)
EPI_list <- read.csv("./Epi_List_2.547_list.txt", header = T)
ID_list <- read.csv("./ID_list_3.1632_list.txt", header = T)
```

##D28

Gene lists of interest: 
-DEGs
```{r}
#HNRNPUdel/+ vs CTRL D28
DEG_D28 <- read.csv("./DEG_ASD12D28_vs_CTRL9D28.csv", sep=,)
DEG_D28sig <- DEG_D28[DEG_D28$baseMean> 20 & DEG_D28$padj < 0.05,]
DEG_D28sig_all <- DEG_D28sig[DEG_D28sig$log2FoldChange< -0.58 | DEG_D28sig$log2FoldChange> 0.58,]
DEG_D28sig_all <- filter(DEG_D28sig, ensembl != "NA")
DEG_D28sig_all <- DEG_D28sig_all[,8]

#siHNRNPU vs siNTC D28
DEG_siD28 <- read.csv("./DEG_si28_NO_outlier.csv", sep=,)
DEG_siD28sig <- DEG_siD28[DEG_siD28$baseMean> 20 & DEG_siD28$padj < 0.05,]
DEG_siD28sig_all <- DEG_siD28sig[DEG_siD28sig$log2FoldChange< -0.58 | DEG_siD28sig$log2FoldChange> 0.58,]
DEG_siD28sig_all <- filter(DEG_siD28sig, ensembl != "NA")
DEG_siD28sig_all <- DEG_siD28sig_all[,8]
```

-Genes in switched compartment 
```{r}
concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D28 <- read_csv("../HiC/HiC_files/concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D28.csv")
concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D28 <- read_csv("../HiC/HiC_files/concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D28.csv")
expressed_FC_GENES_BtoA_switch_regions_D28 <- read_csv("../HiC/HiC_files/expressed_FC_GENES_BtoA_switch_regions_D28.csv")
expressed_FC_GENES_AtoB_switch_regions_D28 <- read_csv("../HiC/HiC_files/expressed_FC_GENES_AtoB_switch_regions_D28.csv")
```

--Joined lists genes in switched compartment: 
```{r}
concordant <- as.data.frame(unique(c(concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D28$ensembl, concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D28$ensembl)))
expressed <- as.data.frame(unique(c(expressed_FC_GENES_AtoB_switch_regions_D28$ensembl, expressed_FC_GENES_BtoA_switch_regions_D28$ensembl)))
colnames(concordant) <- "ensembl"
colnames(expressed) <- "ensembl"
```

Background gene list of HNRNPUdel/+ vs CTRL at  D28

```{r}
all_D28 <- read.csv("./D28_all_expressed_genes.txt", header = T)
backgroundGeneList <-all_D28
```

Enrichment of HNRNPUdel/+ vs CTRL DEGs at D28

```{r}
#SFARI_sel (SFARI scores: S,1,2)
m <- nrow(SFARI_sel_list)
n <- length(setdiff(backgroundGeneList$ensembl,SFARI_sel_list$gene))
q <- length(intersect(DEG_D28sig_all,SFARI_sel_list$gene)) - 1
k <- length(DEG_D28sig_all)
phyper(q,m,n,k, lower.tail = FALSE)

#DD
m <- nrow(DD_list)
n <- length(setdiff(backgroundGeneList$ensembl,DD_list$gene))
q <- length(intersect(DEG_D28sig_all,DD_list$gene)) - 1
k <- length(DEG_D28sig_all)
phyper(q,m,n,k, lower.tail = FALSE)

#EPI
m <- nrow(EPI_list)
n <- length(setdiff(backgroundGeneList$ensembl,EPI_list$gene))
q <- length(intersect(DEG_D28sig_all,EPI_list$gene)) - 1
k <- length(DEG_D28sig_all)
phyper(q,m,n,k, lower.tail = FALSE)

#ID
m <- nrow(ID_list)
n <- length(setdiff(backgroundGeneList$ensembl,ID_list$gene))
q <- length(intersect(DEG_D28sig_all,ID_list$gene)) - 1
k <- length(DEG_D28sig_all)
phyper(q,m,n,k, lower.tail = FALSE)
```

Enrichment of HNRNPUdel/+ vs CTRL genes in switched compartment at D28
```{r}
source("enrichment_phyper.R")

#SFARI sel
CAB <- enrichment_phyper(SFARI_list_NEW, concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D28, backgroundGeneList)
CBA <- enrichment_phyper(SFARI_list_NEW, concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D28, backgroundGeneList)
EAB <- enrichment_phyper(SFARI_list_NEW, expressed_FC_GENES_AtoB_switch_regions_D28, backgroundGeneList)
EBA <- enrichment_phyper(SFARI_list_NEW, expressed_FC_GENES_BtoA_switch_regions_D28, backgroundGeneList)
conc <- enrichment_phyper(SFARI_list_NEW, concordant, backgroundGeneList)
expr <- enrichment_phyper(SFARI_list_NEW, expressed, backgroundGeneList)
SFARI <- rep("SFARI", 3)
variable <- c("p-value", "q", "m")
RES_SFARI <- data.frame(NDD_list = SFARI, variable = variable, ConcordantAtoB = CAB, ConcordantBtoA = CBA, ExpressedAtoB = EAB, ExpressedBtoA = EBA, Concordant = conc, Expressed = expr)

#ID 
CAB <- enrichment_phyper(ID_list, concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D28, backgroundGeneList)
CBA <- enrichment_phyper(ID_list, concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D28, backgroundGeneList)
EAB <- enrichment_phyper(ID_list, expressed_FC_GENES_AtoB_switch_regions_D28, backgroundGeneList)
EBA <- enrichment_phyper(ID_list, expressed_FC_GENES_BtoA_switch_regions_D28, backgroundGeneList)
conc <- enrichment_phyper(ID_list, concordant, backgroundGeneList)
expr <- enrichment_phyper(ID_list, expressed, backgroundGeneList)
ID <- rep("ID", 3)
variable <- c("p-value", "q", "m")
RES_ID <- data.frame(NDD_list = ID, variable = variable, ConcordantAtoB = CAB, ConcordantBtoA = CBA, ExpressedAtoB = EAB, ExpressedBtoA = EBA, Concordant = conc, Expressed = expr)

#EPI
CAB <- enrichment_phyper(EPI_list_NEW, concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D28, backgroundGeneList)
CBA <- enrichment_phyper(EPI_list_NEW, concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D28, backgroundGeneList)
EAB <- enrichment_phyper(EPI_list_NEW, expressed_FC_GENES_AtoB_switch_regions_D28, backgroundGeneList)
EBA <- enrichment_phyper(EPI_list_NEW, expressed_FC_GENES_BtoA_switch_regions_D28, backgroundGeneList)
conc <- enrichment_phyper(EPI_list_NEW, concordant, backgroundGeneList)
expr <- enrichment_phyper(EPI_list_NEW, expressed, backgroundGeneList)
EPI <- rep("EPI", 3)
variable <- c("p-value", "q", "m")
RES_EPI <- data.frame(NDD_list = EPI, variable = variable, ConcordantAtoB = CAB, ConcordantBtoA = CBA, ExpressedAtoB = EAB, ExpressedBtoA = EBA, Concordant = conc, Expressed = expr)

#DD
CAB <- enrichment_phyper(DD_list, concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D28, backgroundGeneList)
CBA <- enrichment_phyper(DD_list, concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D28, backgroundGeneList)
EAB <- enrichment_phyper(DD_list, expressed_FC_GENES_AtoB_switch_regions_D28, backgroundGeneList)
EBA <- enrichment_phyper(DD_list, expressed_FC_GENES_BtoA_switch_regions_D28, backgroundGeneList)
conc <- enrichment_phyper(DD_list, concordant, backgroundGeneList)
expr <- enrichment_phyper(DD_list, expressed, backgroundGeneList)
DD <- rep("DD", 3)
variable <- c("p-value", "q", "m")
RES_DD <- data.frame(NDD_list = DD, variable = variable, ConcordantAtoB = CAB, ConcordantBtoA = CBA, ExpressedAtoB = EAB, ExpressedBtoA = EBA, Concordant = conc, Expressed = expr)
```

```{r}
df_list <- list(RES_SFARI, RES_ID, RES_EPI, RES_DD)
ResultsDF <- df_list %>% reduce(full_join)
write_xlsx(ResultsDF,"../HiCEnrichments.xlsx")
```

FDR corrections: 
```{r}
pvalues <- as.vector(ResultsDF[c(1,4,7,10), 3:8])
values <- c(pvalues[1,], pvalues[2,], pvalues[3,], pvalues[4,])
p.adjust(values, method = "fdr")
```


Background gene list of siHNRNPUdel vs siNTC at  D28
```{r}
all_siD28 <- read.csv("./siD28_all_expressed_genes.csv", header = T, sep=",")
backgroundGeneList <-all_siD28
```


```{r}
#siD28 SFARI_sel
m <- nrow(SFARI_sel_list)
n <- length(setdiff(backgroundGeneList$x,SFARI_sel_list$gene))
q <- length(intersect(DEG_siD28sig_all,SFARI_sel_list$gene)) - 1
k <- length(DEG_siD28sig_all)
phyper(q,m,n,k, lower.tail = FALSE)

#siD28 DD
m <- nrow(DD_list)
n <- length(setdiff(backgroundGeneList$x,DD_list$gene))
q <- length(intersect(DEG_siD28sig_all,DD_list$gene)) - 1
k <- length(DEG_siD28sig_all)
phyper(q,m,n,k, lower.tail = FALSE)

#siD28 EPI
m <- nrow(EPI_list)
n <- length(setdiff(backgroundGeneList$x,EPI_list$gene))
q <- length(intersect(DEG_siD28sig_all,EPI_list$gene)) - 1
k <- length(DEG_siD28sig_all)
phyper(q,m,n,k, lower.tail = FALSE)

#siD28 ID
m <- nrow(ID_list)
n <- length(setdiff(backgroundGeneList$x,ID_list$gene))
q <- length(intersect(DEG_siD28sig_all,ID_list$gene)) - 1
k <- length(DEG_siD28sig_all)
phyper(q,m,n,k, lower.tail = FALSE)


ResultsDF <-read.csv("phyper_D28_results.txt", header=T, sep="\t")       
pvalues <- as.vector(ResultsDF$pvalue)
p.adjust(pvalues, method = "fdr")

```

##D0

Background:
```{r}
D0_all_expressed_genes <- read_excel("./D0_all_expressed_genes.csv")
backgroundGeneList <- as.data.frame(D0_all_expressed_genes$ensembl_gene_id)
```


Gene lists of interest: 
```{r}
concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D0 <- read_csv("/HiC/concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D0.csv")
concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D0 <- read_csv("/HiC/concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D0.csv")
expressed_FC_GENES_AtoB_switch_regions_D0 <- read_csv("/HiC/expressed_FC_GENES_AtoB_switch_regions_D0.csv")
expressed_FC_GENES_BtoA_switch_regions_D0 <- read_csv("/HiC/expressed_FC_GENES_BtoA_switch_regions_D0.csv")
names(concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D0)[names(concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D0) == 'ensembl.1'] <- 'ensembl'
names(concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D0)[names(concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D0) == 'ensembl.1'] <- 'ensembl'
names(expressed_FC_GENES_AtoB_switch_regions_D0)[names(expressed_FC_GENES_AtoB_switch_regions_D0) == 'ensembl.1'] <- 'ensembl'
names(expressed_FC_GENES_BtoA_switch_regions_D0)[names(expressed_FC_GENES_BtoA_switch_regions_D0) == 'ensembl.1'] <- 'ensembl'

#Joined lists: 

concordant <- as.data.frame(unique(c(concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D0$ensembl, concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D0$ensembl)))
expressed <- as.data.frame(unique(c(expressed_FC_GENES_AtoB_switch_regions_D0$ensembl, expressed_FC_GENES_BtoA_switch_regions_D0$ensembl)))
colnames(concordant) <- "ensembl"
colnames(expressed) <- "ensembl"
```

SFARI:

```{r}
CAB <- enrichment_phyper(SFARI_list_NEW, concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D0, backgroundGeneList)
CBA <- enrichment_phyper(SFARI_list_NEW, concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D0, backgroundGeneList)
EAB <- enrichment_phyper(SFARI_list_NEW, expressed_FC_GENES_AtoB_switch_regions_D0, backgroundGeneList)
EBA <- enrichment_phyper(SFARI_list_NEW, expressed_FC_GENES_BtoA_switch_regions_D0, backgroundGeneList)
conc <- enrichment_phyper(SFARI_list_NEW, concordant, backgroundGeneList)
expr <- enrichment_phyper(SFARI_list_NEW, expressed, backgroundGeneList)
SFARI <- rep("SFARI", 3)
variable <- c("p-value", "q", "m")
RES_SFARI <- data.frame(NDD_list = SFARI, variable = variable, ConcordantAtoB = CAB, ConcordantBtoA = CBA, ExpressedAtoB = EAB, ExpressedBtoA = EBA, Concordant = conc, Expressed = expr)
```

ID: 

```{r}
CAB <- enrichment_phyper(ID_list, concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D0, backgroundGeneList)
CBA <- enrichment_phyper(ID_list, concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D0, backgroundGeneList)
EAB <- enrichment_phyper(ID_list, expressed_FC_GENES_AtoB_switch_regions_D0, backgroundGeneList)
EBA <- enrichment_phyper(ID_list, expressed_FC_GENES_BtoA_switch_regions_D0, backgroundGeneList)
conc <- enrichment_phyper(ID_list, concordant, backgroundGeneList)
expr <- enrichment_phyper(ID_list, expressed, backgroundGeneList)
ID <- rep("ID", 3)
variable <- c("p-value", "q", "m")
RES_ID <- data.frame(NDD_list = ID, variable = variable, ConcordantAtoB = CAB, ConcordantBtoA = CBA, ExpressedAtoB = EAB, ExpressedBtoA = EBA, Concordant = conc, Expressed = expr)
```

EPI: 

```{r}
CAB <- enrichment_phyper(EPI_list_NEW, concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D0, backgroundGeneList)
CBA <- enrichment_phyper(EPI_list_NEW, concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D0, backgroundGeneList)
EAB <- enrichment_phyper(EPI_list_NEW, expressed_FC_GENES_AtoB_switch_regions_D0, backgroundGeneList)
EBA <- enrichment_phyper(EPI_list_NEW, expressed_FC_GENES_BtoA_switch_regions_D0, backgroundGeneList)
conc <- enrichment_phyper(EPI_list_NEW, concordant, backgroundGeneList)
expr <- enrichment_phyper(EPI_list_NEW, expressed, backgroundGeneList)
EPI <- rep("EPI", 3)
variable <- c("p-value", "q", "m")
RES_EPI <- data.frame(NDD_list = EPI, variable = variable, ConcordantAtoB = CAB, ConcordantBtoA = CBA, ExpressedAtoB = EAB, ExpressedBtoA = EBA, Concordant = conc, Expressed = expr)
```


DD: 
```{r}
CAB <- enrichment_phyper(DD_list, concordantFC_expressed_DEG_GENES_AtoB_switch_regions_D0, backgroundGeneList)
CBA <- enrichment_phyper(DD_list, concordantFC_expressed_DEG_GENES_BtoA_switch_regions_D0, backgroundGeneList)
EAB <- enrichment_phyper(DD_list, expressed_FC_GENES_AtoB_switch_regions_D0, backgroundGeneList)
EBA <- enrichment_phyper(DD_list, expressed_FC_GENES_BtoA_switch_regions_D0, backgroundGeneList)
conc <- enrichment_phyper(DD_list, concordant, backgroundGeneList)
expr <- enrichment_phyper(DD_list, expressed, backgroundGeneList)
DD <- rep("DD", 3)
variable <- c("p-value", "q", "m")
RES_DD <- data.frame(NDD_list = DD, variable = variable, ConcordantAtoB = CAB, ConcordantBtoA = CBA, ExpressedAtoB = EAB, ExpressedBtoA = EBA, Concordant = conc, Expressed = expr)
```


```{r}
df_list <- list(RES_SFARI, RES_ID, RES_EPI, RES_DD)
ResultsDF <- df_list %>% reduce(full_join)
```

```{r}
write_xlsx(ResultsDF,"../HiCEnrichmentsD0.xlsx")
```

FDR corrections: 

```{r}
pvalues <- as.vector(ResultsDF[c(1,4,7,10), 3:8])
values <- c(pvalues[1,], pvalues[2,], pvalues[3,], pvalues[4,])
p.adjust(values, method = "fdr")
```
